---
title: "2: Physical Properties of Lakes"
author: "Water Data Analytics | Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Investigate the concepts of lake stratification and mixing by analyzing monitoring data
2. Apply data analytics skills to applied questions about physical properties of lakes
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion

What are the physical properties of lakes?

## Session Set Up
```{r, message = FALSE}
# Check working directory (should be project file location)
getwd()

# install.packages("tidyverse")
# install.packages("lubridate")
# install.packages("rLakeAnalyzer")

# load packages
library(tidyverse)
library(lubridate)
library(rLakeAnalyzer)

# Load data
NTLdata <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
#here package also works with relative file paths 
#North Temperate Lakes long term ecological monitoring data 

# set ggplot theme
theme_set(theme_classic())
#have a theme in ggplot so things look consistent across a project 
```

## Data Wrangling and Exploration

### Investigate structure of dataset

Common steps/checks for data wrangling: 

* Date formatting
* Addressing NAs
* Subsetting and filtering 

```{r}
# Is the date column perceived as a date by R?
class(NTLdata$sampledate)
#no it is perceived as a character 
NTLdata$sampledate <- as.Date(NTLdata$sampledate, format = "%m/%d/%y")
#as.Date changes from character to date, and we provide the existing date format (lookup date codes for upper/lower case)
#help(as.Date)
class(NTLdata$sampledate) #now date is recognized


# What does this column look like now?
View(NTLdata)
#1984-05-27 date structure

# Remove rows that contain NAs in temperature column
dim(NTLdata) #dim is the dimensions 
NTLdata <- NTLdata %>% #read pipe as "then"
  drop_na(temperature_C) #got rid of all NA values in temp column
dim(NTLdata)  
#here we are using dplyr, so we don't need to use $ because dplyr knows within the pipe what dataset we are working with
#alternative code for the above
# NTLData <- drop_na(NTLdata, temperature_C)
#however pipes are easy to add new things on top of 

# How many observations are there for each lake?
summary(NTLdata$lakename) #this treats it as a character
summary(as.factor(NTLdata$lakename)) #here instead it is a factor of the column
#this also tells us how many temperature measurements we have
#Peter and Paul lake have the longest temperature data collection

# Let's choose the two lakes with the most data
NTLdata_PeterPaul <- NTLdata %>%
  filter(lakename %in% c("Paul Lake", "Peter Lake")) #c is concatenate
# %in% reads as "includes"

# What is another way to use the filter command to get to the same result?
NTLdata_PeterPaul <- NTLdata %>%
  filter(lakename == "Paul Lake" | lakename == "Peter Lake") 
# | means "or"

# Make two data frames, one for each lake
Pauldata <- NTLdata %>%
  filter(lakename == "Paul Lake")
Peterdata <- NTLdata %>%
  filter(lakename == "Peter Lake")

# How long did the monitoring last?
min(Pauldata$sampledate)
max(Pauldata$sampledate)
min(Peterdata$sampledate)
max(Peterdata$sampledate)
#1984-2016

# Which depths are sampled in each lake? (in meters)
unique(Pauldata$depth) 
unique(Peterdata$depth)
# Why didn't we use the "summary" function here?
#summary gives min and max but doesn't give discrete sample locations 

#20 meters is either a typo or a one time sample
# QA the one data point at 20 m depth. 
Pauldata <- Pauldata %>%
  filter(depth < 20)

```

### Exploratory data visualization

Let's make a plot of temperatures by depth. There are a lot of points, so adding a 50 % transparency to the points helps us see where points are densely clustered together.

Let's also try using the traditional limnological graph type, with depth on the y axis in reverse, to simulate a cross section of a lake. When and where do we usually observe high and low temperatures?
```{r}
ggplot(Pauldata, aes(x = depth, y = temperature_C)) + 
  geom_point(alpha = 0.5) + #alpha is transparency 
  labs(y = expression("Temperature "(degree*C)), x = "Depth (m)")

ggplot(Pauldata, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point(alpha = 0.5) +
  scale_y_reverse() +
  scale_color_viridis_c() +
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)", color = "DOY")+ #DOY is day of year
  theme_classic()
#the cold yellow points are very late in the year 
#in the data we can see this is the only year that they sampled in november
#in that november sample, the lake is thermally mixed, you might expect other variables to mixed as well 
#the summer (green points) has a warm top(epilimion) and cold bottom(hypolimnion), is very stratified


#limnology has independent on y and dependent on x
#limnologists like to also flip it to look at it as a water column as in the second plot

```

How do temperatures at the surface compare to temperatures at the mid-depths and at the bottom?
#Lake Types: (mictic means mixing)
#dimictic lakes mix twice in the year, spring and fall will have mixed water columns
  #winter will have 0 degrees at top and 4 degrees at the bottom
  #summer will have colder at the bottom and warmer at the top 
#monomictic lakes mix one time. The thermal mass of the water column have different densities and resist mixing. It is more likely to have stronger stratification when the air is very warm and the bottom of the water column remains cool, during the summer is when they have the strongest stratification. 
#polymictic mixes several times per year - mixing happens with strong wind events or weather patterns as opposed to seasonal. Shallower lakes are more likely to be polymictic. They tend to have narrower thermal profiles. 
#amictic - these lakes are always stratified and do not mix. They tend to be very deep and sheltered from weather patterns. 


### Graphing seasonal water profiles

Lakes in the North Temperate Lakes LTER are dimictic, meaning they mix fully twice per year. When the lakes are not mixed (stratified), the top layer (epilimnion) and the bottom layer (hypolimnion) of the lake are different temperatures.

For background information on lake stratification, see this [stratification summary article from the North American Lake Management Society](https://www.nalms.org/secchidipin/monitoring-methods/temperature-and-oxygen/).


In 1993, Paul Lake was sampled from May 20 through November 2. Let's look at temperature profiles of the lake across the year and also explore how this impacts dissolved oxygen.

```{r}
ggplot(subset(Pauldata, year4 == 1993), 
       aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point(alpha = 0.5) +
  scale_y_reverse() +
  scale_color_viridis_c() +
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)", color = "DOY")+
  theme_classic()

# Now, make a plot as above but with dissolved oxygen saturation instead of temperature.
ggplot(subset(Pauldata, year4 == 1993), 
       aes(x = dissolvedOxygen, y = depth, color = daynum)) +
  geom_point(alpha = 0.5) +
  scale_y_reverse() +
  scale_color_viridis_c() +
  labs(x = "DO mg/L", y = "Depth (m)", color = "DOY")

```

Why does mixing occur in the spring and fall? What are the mechanisms that make this possible?
During stratified conditions, 
DO concentrations are higher in the epilimnion (top) which is where wind induced mixing occurs increasing DO. Plants and algae might also contribute to more DO in the upper water column. 
There is a slight bulge in DO right before the thermocline, which is where the algae stay because this is the depth where nutrients concentrate. Respiration in the hypolimnion, dead and decaying algae sink down and phytoplankton hang out right where there is enoguh light and a peak in nutrients "deep chlorophyl maximum". There are also higher concentrations of CO2 and methane (products of anaerobic respiration) in the hypolimnion. 


### Calculating thermoclines

[rLakeAnalyzer documentation](https://cran.r-project.org/web/packages/rLakeAnalyzer/rLakeAnalyzer.pdf)

```{r}
Pauldata_thermo <- Pauldata %>% #creating a new dataset 
  group_by(year4, daynum, sampledate) %>% #grouping dataframe by dates (keeping year and daynum)
  # calculate thermoclines based on temperature profiles. 
  # seasonal = FALSE calculates the thermocline as the maximum density gradient
  # seasonal = TRUE calculates the thermocline as the deepest density gradient
  #think about what would make more sense in specific contexts 
  summarise(thermocline = thermo.depth(wtr = temperature_C, depths = depth, seasonal = FALSE)) %>%
  # remove all thermoclines within 1 m of the surface. these can represent transient stratification.
  #naming new column thermocline
  #thermo.depth function takes temperature profile and associated depths 
  filter(thermocline > 1) #taking out transient stratificiation and removing blanks 

View(Pauldata_thermo)

ggplot(Pauldata_thermo, aes(x = daynum, y = thermocline)) +
  geom_point() 
  
```

How does thermocline depth change over the course of the year? When do we observe the most variability in thermocline depth?
An overall increase in thermocline depth over the course of the year is emerging (thermocline deepens as the year goes on). There is more variability earlier in the year and it tends to be more consistent later in the year. (Day zero in limnology is when stratification starts to take place). 


### Climate change

Increases in global temperature are predicted to cause several changes to lake thermal conditions, including: 

* Increases in surface temperature
* Increases in surface minimum temperature
* Increases in extent of stratification
* Increases in length of stratification
* Decreases in ice cover - potential for lakes to be switching from dimictic to monomictic 

Several studies have addressed this topic, using long-term and spatially diverse datasets: 

https://link.springer.com/article/10.1007/s10584-015-1326-1?sa_campaign=email/event/articleAuthor/onlineFirst&error=cookies_not_supported&error=cookies_not_supported&code=2b415e25-de4c-452f-bd02-2cceae08b7a3&code=e63aabb9-76d3-4e49-b36c-e591007a9e9c

http://hpkx.cnjournals.com/uploadfile/news_images/hpkx/2020-07-15/10.1038-s43017-020-0067-5.pdf

https://agupubs.onlinelibrary.wiley.com/doi/pdf/10.1002/2015GL064097

https://link.springer.com/article/10.1007/s10584-019-02465-y

Let's explore how these changes might manifest in Paul Lake. Here we have over 30 years of data. Can we see changes in climate in this dataset? 
```{r}
ggplot(Pauldata_thermo, aes(x = daynum, y = thermocline, color = year4)) +
  geom_point() +
  scale_color_viridis_c()

# exercise: put the legend on top and relabel the aesthetics

#If we were seeing increases in extent and length of stratification, we might expect thermocline depth to narrow, and to be deeper earlier in the season
#First day of stratification might happen earlier

#plotting surface temperatures
ggplot(subset(Pauldata, depth == 1),
       aes(x = daynum, y = temperature_C, color = as.factor(year4))) + # year as a factor here 
  geom_point(alpha = 0.5) +
  geom_line() +
  scale_color_viridis_d()
# exercise: relabel the aesthetics

```

## Closing Discussion

What are the main concepts you learned about the physical properties of lakes today? What was the evidence for these concepts in the dataset?
