---
title: "4: Physical Properties of Rivers"
author: "Water Data Analytics | Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Compute recurrence intervals for stream discharge 
2. Analyze the effects of watershed disturbance on recurrence intervals and interpret results against the concept of stationarity
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
How is climate change impacting discharge in streams and rivers? What are the new and ongoing challenges faced by these impacts in watersheds?
-Higher flow variability (snowmelt changing)
-Shifts in precipitation type
-Greater evapotranspiration (less discharge entering streams and rivers)

## Session Set Up

```{r, message = FALSE}
getwd()

# install.packages("lfstat")

library(tidyverse)
library(dataRetrieval)
library(lubridate)
library(lfstat) #used for low flow statistics


theme_set(theme_classic())
```

## Recurrence Intervals and Exceededence Probability

A **recurrence interval** is the past recurrence of an event, in this case a peak annual discharge measurement of a given magnitude. The value of a recurrence interval corresponds to the average number of years between discharge of a given magnitude. Typically the minimum amount of years required to construct a recurrence interval is 10, but 30 is more robust. A recurrence interval, T, is calculated as: 

$T = (n+1)/m$ #this will show as an equation if the document is knit

where n is the number of years and m is the ranking of an event within the observed period. We add one to n because we are computing the recurrence interval for a discharge event of a given magnitude *or greater*. 

Similarly, we can calculate an **exceedence probability**, or the probability of encountering a discharge event of a given magnitude or greater in any given year: 

$P = 1/T$

This is where the terms "100-year flood" and similar are derived. Remember this is a probability based on past occurrence, not an accurate forecast of how often we will see that event happening. When current patterns of discharge differ from past patterns, we observe **nonstationary** behavior. Nonstationarity results in events that occur more or less frequency than predicted based on the exceedence probability.

### Has Eno River dicharge displayed stationary behavior over the period of record? Stationary means that the probability of events of a certain magnitude happening has been the same over time - variability is not trending upward or downward. 

Let's import discharge data for the Eno River near Durham for all available dates.

```{r}
EnoDischarge <- readNWISdv(siteNumbers = "02085070",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "", 
                     endDate = "2021-09-30")
#getting earliest period of record start, ending at the end of the most recent water year

#relabeling column names
names(EnoDischarge)[4:5] <- c("Discharge", "Approval.Code")

#viewing dataset attributes
attr(EnoDischarge, "variableInfo") 
attr(EnoDischarge, "siteInfo")
  
# Build a ggplot for discharge over time
ggplot(EnoDischarge, aes(x = Date, y = Discharge)) +
  geom_line() +
  labs(x = "Year", y = "Discharge (cfs)")
```

We can then compute recurrence intervals based on the first 30 years of data.

```{r}  
class(EnoDischarge$WaterYear)

#generating new columns
EnoDischarge <- EnoDischarge %>%
  mutate(Year = year(Date), 
         WaterYear = water_year(Date, origin = "usgs")) %>% #using USGS definition of water year 
  filter(WaterYear != "1963") #excluding this water year because this was an incomplete water year from the data
#in quotes because R thinks it is a factor

# Water Year is a factor. We want to re-classify as numeric, have to classify it as a character first then define the water year
EnoDischarge$WaterYear <- as.numeric(as.character(EnoDischarge$WaterYear))

#Calculating recurrence intervals! kind of a pivot table
EnoRecurrence <- 
  EnoDischarge %>%
  filter(WaterYear < 1994) %>%
  group_by(WaterYear) %>%
  summarise(PeakDischarge = max(Discharge)) %>% 
  mutate(Rank = rank(-PeakDischarge), #giving number 30 to the lowest value
         RecurrenceInterval = (length(WaterYear) + 1)/Rank, 
         Probability = 1/RecurrenceInterval)
#year rank number 1 is going to have a recurrence interval of 31

#plot the timeseries 
ggplot(EnoRecurrence, aes(x = WaterYear, y = PeakDischarge)) +
  geom_bar(stat = "identity") +
  labs(x = "Year", y = "Peak Discharge (cfs)")

```

Let's display and model the relationship between peak annual disharge and recurrence interval. We can use the statistical model to compute discharge for recurrence intervals that occur above the 30-year mark.
```{r}
#plotting recurrence interval against peak discharge
ggplot(EnoRecurrence, aes(x = RecurrenceInterval, y = PeakDischarge)) +
  geom_point() +
  scale_x_log10() + #note log scale to linearize data 
  geom_smooth(method = "lm", color = "black", se = FALSE) #linear model geomsmooth
# removing se = FALSE includes standard error bars

Eno.RImodel <- lm(data = EnoRecurrence, PeakDischarge ~ log10(RecurrenceInterval)) #linear regression model! 
summary(Eno.RImodel) #tells us the equation of this line and R^2 value

#now that we have our model, we can plug in our recurrence intervals
#this will predict what the discharge will be for those years
#What is the discharge for a 100-year flood in this system? a 500-year flood?
Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log10(100) 
Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log10(500)
```

What if we were to build a recurrence interval model for the most recent 30 years? How would this compare to the early period recurrence interval?
```{r}
#has the eno river displayed stationarity? 

#same as the code above for another set of ears 
EnoRecurrence.Late <- 
  EnoDischarge %>%
  filter(WaterYear >= 1992) %>%
  group_by(WaterYear) %>%
  summarise(PeakDischarge = max(Discharge)) %>% 
  mutate(Rank = rank(-PeakDischarge), 
         RecurrenceInterval = (length(WaterYear) + 1)/Rank, 
         Probability = 1/RecurrenceInterval)

#adding the second curve to the plot to compare the two 
ggplot(EnoRecurrence, aes(x = RecurrenceInterval, y = PeakDischarge)) +
  geom_point() +
  geom_point(data = EnoRecurrence.Late, color = "#02818a",
             aes(x = RecurrenceInterval, y = PeakDischarge)) +
  scale_x_log10()

#how does it hold up with the model? 
Eno.RImodel.Late <- lm(data = EnoRecurrence.Late, PeakDischarge ~ log10(RecurrenceInterval))

summary(Eno.RImodel.Late)
#comparing the coefficients of the two models 
#how does the slope change? 
Eno.RImodel.Late$coefficients
Eno.RImodel$coefficients

#how to the 100 and 500 year flood events change?
Eno.RImodel.Late$coefficients[1] + Eno.RImodel.Late$coefficients[2]*log10(100)
Eno.RImodel.Late$coefficients[1] + Eno.RImodel.Late$coefficients[2]*log10(500)

Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log10(100)
Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log10(500)
#use this information to think about how hydrology affects the stationarity of a model! 
```

What differences did you see for the recurrence intervals built under different periods of record? How would your prediction of flood events differ if you were to use these models for forecasting purposes? 

>the late period of record had a much more massive 100 and 500 year flood events!

What would you recommend for a watershed manager seeking to build the most accurate recurrence interval model for the Eno River?

> use the more recent period of recurrence for management decisions


### Examining the effects of urbanization on discharge

Salado Creek is located in San Antonio, Texas, an area that has been rapidly urbanizing over the course of the last several decades (http://worldpopulationreview.com/us-cities/san-antonio-population/#byPopulation). Is this system exhibiting stationarity?

```{r}
# Import data
SaladoDischarge <- readNWISdv(siteNumbers = "08178700",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "")
names(SaladoDischarge)[4:5] <- c("Discharge", "Approval.Code")
attr(SaladoDischarge, "siteInfo")
  
ggplot(SaladoDischarge, aes(x = Date, y = Discharge)) +
  geom_line() +
  labs(x = "Year", y = "Discharge (cfs)")

```

### Examining the effects of dam construction on recurrence intervals

The stream gage in the Green River near Auburn, Washington, is located directly downstream of the Howard A. Hanson Dam. The dam was built in 1961 for flood control purposes, and the reservoir now provides water supply to the city of Tacoma. How have peak discharges changed since the construction of the dam?

```{r}
GreenDischarge <- readNWISdv(siteNumbers = "12113000",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "")
names(GreenDischarge)[4:5] <- c("Discharge", "Approval.Code")
attr(GreenDischarge, "siteInfo")
  
ggplot(GreenDischarge, aes(x = Date, y = Discharge)) +
  geom_line() +
  labs(x = "Year", y = "Discharge (cfs)")
#dams modulate the peakiness of discharge
#always look into the history of your system as you perceive change over time

```

## Bonus content: Flow Duration Curves and Low Flow Statistics

Flow-duration curves can be generated from daily discharge data, similar to how we calculated recurrence intervals for annual data. 

$P = 100*(m/(n+1))$

where P is the exceedance probability, m is the ranking of all daily mean flows in the period of record (at least 10 years), and n is the total number of daily mean flows. 

We focused today on recurrence intervals, which use peak flow statistics. On the other end of the discharge gradient are low flow statistics, most commonly estimated by 7Q2 and 7Q10 metrics (7-day, 2-year and 10-year annual low flow statistics). These can be used to evaluate drought conditions and are another metric for evaluating stationarity in rivers and streams. 

See the USGS description of these statistics here:(Calculating Flow-Duration and Low-Flow Frequency Statistics at Streamflow-Gaging Stations)[https://pubs.usgs.gov/sir/2008/5126/section3.html]
