---
title: "Assignment 4: Physical Properties of Rivers"
author: "Jackie Van Der Hout"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on the physical properties of rivers. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, check your PDF against the key and then submit your assignment completion survey at https://forms.gle/futQwtCsyYsZG9nCA

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-02-15

## Setup and Data Processing

1. Verify your working directory is set to the R project file. Load the tidyverse, cowplot, dataRetrieval, lubridate, lfstat, and EcoHydRology packages. Set your ggplot theme (can be theme_classic or something else).
2. Acquire daily mean discharge data for the Bitterroot River in Montana (USGS gage 12344000) and the Nehalem River in Oregon (USGS gage 14299800). Collect the 10 most recent complete water years.
3. Add columns in the dataset for water year, baseflow, and stormflow. Feel free to use either baseflow separation function we used in class. 
4. Calculate annual total discharge, annual baseflow, and annual proportion baseflow for the two sites.

```{r}
#1
getwd()
library(EcoHydRology)
library(tidyverse)
library(cowplot)
library(dataRetrieval)
library(lubridate)
library(lfstat)

theme_set(theme_light())

#2
BitterrootDischarge <- readNWISdv(siteNumbers = "12344000",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2012-10-01", 
                     endDate = "2021-09-30")

names(BitterrootDischarge)[4:5] <- c("Discharge", "Approval.Code")
attr(BitterrootDischarge, "variableInfo")
attr(BitterrootDischarge, "siteInfo")

NehalemDischarge <- readNWISdv(siteNumbers = "14299800",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2012-10-01", 
                     endDate = "2021-09-30")

names(NehalemDischarge)[4:5] <- c("Discharge", "Approval.Code")
attr(NehalemDischarge, "variableInfo")
attr(NehalemDischarge, "siteInfo")

#3

BitterrootDischarge <- BitterrootDischarge %>%
  mutate(Baseflow_lfstat = baseflow(Discharge), 
         Stormflow_lfstat = Discharge - Baseflow_lfstat,
         Year = year(Date), 
         WaterYear = water_year(Date))

NehalemDischarge <- NehalemDischarge %>%
  mutate(Baseflow_lfstat = baseflow(Discharge), 
         Stormflow_lfstat = Discharge - Baseflow_lfstat,
         Year = year(Date), 
         WaterYear = water_year(Date))

BitterrootDischarge$WaterYear <- as.numeric(as.character(BitterrootDischarge$WaterYear))
NehalemDischarge$WaterYear <- as.numeric(as.character(NehalemDischarge$WaterYear))
```

## Analyze seasonal patterns in discharge

5. For both sites, create a graph displaying discharge and baseflow by date. Adjust axis labels accordingly.
6. For both sites, create a graph displaying annual total discharge and annual baseflow across years, and a second graph displaying the proportion baseflow across years (adjust axis labels accordingly). Plot these graphs on top of one another using `plot_grid`. Remember to align the axes!

```{r}
#5
BitterrootPlot <- ggplot(BitterrootDischarge, aes(x = Date, y = Discharge)) +
  geom_line(color = "blue", alpha = 0.5) +
  geom_line(aes(y = Baseflow_lfstat)) +
  scale_y_log10() + 
  labs(x = "Year", y = expression("Discharge (ft"^3*"/s)"), title = "Bitterroot River")
BitterrootPlot

NehalemPlot <- ggplot(NehalemDischarge, aes(x = Date, y = Discharge)) +
  geom_line(color = "blue", alpha = 0.5) +
  geom_line(aes(y = Baseflow_lfstat)) +
  scale_y_log10() + 
  labs(x = "Year", y = expression("Discharge (ft"^3*"/s)"), title = "Nehalem River")
NehalemPlot

#6a: Bitterroot

#had to look in guide for this one
BitterrootDischarge_summary <- BitterrootDischarge %>%
  group_by(WaterYear) %>%
  summarise(Discharge.acft.yr = sum(Discharge)*723.968, #cfs to acre feet/year 
            Baseflow.acft.yr = sum(Baseflow_lfstat, na.rm = TRUE)*723.968,
            Stormflow.acft.yr = sum(Stormflow_lfstat, na.rm = TRUE)*723.968,
            Prop.bf = Baseflow.acft.yr/Discharge.acft.yr,
            Prop.sf = Stormflow.acft.yr/Discharge.acft.yr) %>%
 mutate_if(is.numeric, round, 2)

class(BitterrootDischarge_summary$WaterYear) #reclassify
BitterrootDischarge_summary$WaterYear <- as.numeric(as.character(BitterrootDischarge_summary$WaterYear))

AnnualBitterroot <- ggplot(BitterrootDischarge_summary, aes(x = WaterYear, y = Discharge.acft.yr))+
  geom_line()+
  geom_line(aes(y = Baseflow.acft.yr), lty = 2)+
  scale_x_continuous(breaks = c(2012, 2014, 2016, 2018, 2020))+
  labs(x = "Water Year", y = "Discharge (Acre Feet/Year)")

PropBitterroot <- ggplot(BitterrootDischarge_summary, aes(x = WaterYear, y = Prop.bf)) +
  geom_col()+
  labs(x = "Water Year", y = "Baseflow Proportion", title = "")

plot_grid(AnnualBitterroot, PropBitterroot)

#6b: Nehalem
NehalemDischarge_summary <- NehalemDischarge %>%
  group_by(WaterYear) %>%
  summarise(Discharge.acft.yr = sum(Discharge)*723.968, #cfs to acre feet/year 
            Baseflow.acft.yr = sum(Baseflow_lfstat, na.rm = TRUE)*723.968,
            Stormflow.acft.yr = sum(Stormflow_lfstat, na.rm = TRUE)*723.968,
            Prop.bf = Baseflow.acft.yr/Discharge.acft.yr,
            Prop.sf = Stormflow.acft.yr/Discharge.acft.yr) %>%
 mutate_if(is.numeric, round, 2)

class(NehalemDischarge_summary$WaterYear) #reclassify
NehalemDischarge_summary$WaterYear <- as.numeric(as.character(NehalemDischarge_summary$WaterYear))

AnnualNehalem <- ggplot(NehalemDischarge_summary, aes(x = WaterYear, y = Discharge.acft.yr))+
  geom_line()+
  geom_line(aes(y = Baseflow.acft.yr), lty = 2)+
  scale_x_continuous(breaks = c(2012, 2014, 2016, 2018, 2020))+
  labs(x = "Water Year", y = "Discharge (Acre Feet/Year)", title = "")

PropNehalem <- ggplot(NehalemDischarge_summary, aes(x = WaterYear, y = Prop.bf)) +
  geom_col()+
  labs(x = "Water Year", y = "Baseflow Proportion")

plot_grid(AnnualNehalem, PropNehalem)

```

7. How do these rivers differ in their discharge and baseflow, both within and across years? How would you attribute these patterns to the climatic conditions in these locations?

> The Bitterroot river has peakier baseflows, with steep inclines and declines. The Nehalem river has lower overall baseflows but much peakier storm events. This indicates that the Bitterroot is fed more by snow melt while the Nehalem river is fed by seasonal storm events. Both have strong seasonal patterns, but the Bitterroot river seems to have more flow contributions from summer storms where as the Nehalem river has very little summer flow contributions. The Bitterroot river has seen some variblity with overall discharge and baseflow conditions. Across the years, the Nehalem River has seen low baseflows and droughtline conditions in 2013 and more stable discharge quantities since then, with some variability. It is possible that long-standing western drought conditions have affected the watersheds of both of these rivers. 
