---
title: "Assignment 8: Time Series Analysis"
author: "Jackie Van Der Hout"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on time series analysis

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, check your PDF against the key and then submit your assignment completion survey at https://forms.gle/dKEutwXiFewkSTwN9

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-03-29

## Setup

1. Verify your working directory is set to the R project file. Load the tidyverse, lubridate, trend, forecast, and dataRetrieval packages. Set your ggplot theme (can be theme_classic or something else).


```{r setup}
getwd()
library(tidyverse)
library(lubridate)
library(trend)
library(forecast)
library(dataRetrieval)
theme_set(theme_light())
```

## Data Import and Processing

2. Import discharge data (parameter 00060) from Clear Creek, Colorado (site 06719505) from the start of 1975 through the end of 2021.
3. Graph Clear Creek discharge over time. 
4. Create a new data frame with the sum of monthly discharge in acre-feet per month.

```{r}
#2
ClearFlow <- readNWISdv(siteNumbers = "06719505",
                     parameterCd = "00060", 
                     startDate = "1975-01-01", 
                     endDate = "2021-01-01")

names(ClearFlow)[4:5] <- c("Discharge", "Approval.Code")
sum(is.na(ClearFlow)) #no NA values, no need to interpolate or otherwise close gaps 

#3
ggplot(ClearFlow, aes(x = Date, y = Discharge)) +
  geom_line() +
  labs(x = "Date", y = "Discharge (cfs)")

#4
ClearFlowSum <- ClearFlow %>% 
  mutate(Year = year(Date),
        Month = month(Date)) %>% 
  group_by(Year, Month) %>% 
  summarise(Discharge = sum(Discharge)*1.98347)

```


## Time Series Decomposition
5. Create a time series of discharge from the monthly data frame. Make sure to add start and end dates like we did in class.
6. Decompose the time series using the `stl` function.
7. Visualize the decomposed time series.

```{r}
#5
ClearFlow_timeseries <- ts(ClearFlowSum$Discharge, frequency = 12,
                           start = c(1975, 01, 01), end = c(2021, 01, 01))

#6
ClearFlow_decomposed <- stl(ClearFlow_timeseries, s.window = "periodic")

#7
plot(ClearFlow_decomposed)
```

8. How do the seasonal and trend components of the decomposition compare to the Neuse River discharge dataset? 

> Seasonal: Clear Creek appears to have a much stronger seasonal flow pattern than the Neuse river does, with a steeper slope between seasons indicating more a difference in flow. This is hard to detect by sight but is clear upon examination of the scales of the axes between the two plots. 

> Trend: While both trendlines seem to indicate a good amount of variability over time, Clear Creek seems to be experiencing greater variation in trends in both positive and negative directions, as is also apparent when comparing the scale of the trend axes. 

## Trend Analysis
Research question: Has there been a monotonic trend in discharge in Clear Creek over the period of study?

9. Run a Seasonal Mann-Kendall test on the monthly discharge data. Inspect the overall trend and the monthly trends. 

```{r}
ClearFlow_trend <- smk.test(ClearFlow_timeseries)

# Inspect results
ClearFlow_trend
summary(ClearFlow_trend)
#positive z score says that spacing between points is in a positive direction over time 


```

10. Is there an overall monotonic trend in discharge over time? Are there monthly trends over time? If so, are they positive or negative?

> Here the SMK test is not showing a clear trend over time. While there is a slight significant positive trend in flow in April, possibly indicating earlier snowmelt times, overall variability between positive and negative means that the trend is not monotonic. 


## Forecasting
Research question: can we predict discharge in Clear Creek moving into the future?

11. Run the auto.arima function on the Clear Creek time series to search for the best fit. Create an object that defines the best fit model. 
12. Make a prediction into the future and plot the future predictions. 

```{r}
auto.arima(ClearFlow_timeseries, trace = TRUE)

#says with drift, is there anything I need to change? 
fit <- arima(ClearFlow_timeseries, c(1,0,0),seasonal = list(order = c(1,1,0), period = 12))

Clearprediction <- predict(fit, n.ahead = 10*12) 

# plot future predictions  
ts.plot(ClearFlow_timeseries, Clearprediction$pred, lty = c(1, 3))

```

13. How did the forecasting for Clear Creek compare to the Neuse River?

> The Neuse river forecasting decomposed into a single value over time, while the Clear Creek forecasting preserved the seasonal nature of the flow patterns into the future, though it may not have captured the increase in variability, both positive and negative, that the flow regime has experienced. 